---
title: "Reproducible Research Project 1"
author: "Jim Staib"
date: "January 27, 2017"
output: 
  html_document:
    keep_md: true 
---

```{r setoptions, echo=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE, results = "asis")
```
# Loading and preprocessing the data
```{r loaddata}
activity <- read.csv("activity.csv")
```

#What is mean total number of steps taken per day?

```{r meanperday}
StepsPerDay <- with(activity, tapply(steps,date, sum, na.rm=T))
hist(StepsPerDay)
meanSteps <- as.integer(mean(StepsPerDay))
medianSteps <- as.integer(median(StepsPerDay))
```

Mean Steps Per Day is `r meanSteps`.   Median Steps Per Day is `r medianSteps`.

#What is the average daily activity pattern?

```{r avgdailyactivity}
StepsPerInterval <- with(activity, tapply(steps,interval, sum, na.rm=T))
plot(names(StepsPerInterval), StepsPerInterval, type="l", xlab= "Time of Day", ylab= "Steps", col="green" , lwd=2)
maxInterval <- names(which.max(StepsPerInterval))
```

The 5 minute interval with the maximum average number of steps across all days is `r maxInterval`.

#Imputing missing values

To impute missing values, this code uses plyr's ddply to transform the activity data frame using a custom "impute.mean" function.   Because the results are out of order, the dataframe is resorted using to match the original sort order

```{r imputemissing}
#sum missing number of steps
missingSteps <- sum(is.na(activity$steps)) 
#replace na with mean
library(plyr)
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
actImpute <- ddply(activity, ~ interval, transform, steps = impute.mean(steps))
actImpute <- actImpute[order(actImpute$date,actImpute$interval), ] 
StepsPerDayImpute <- with(actImpute, tapply(steps,date, sum, na.rm=T))
hist(StepsPerDayImpute)
meanStepsImpute <- as.integer(mean(StepsPerDayImpute))
medianStepsImpute <- as.integer(median(StepsPerDayImpute))
```

Mean Steps Per Day using imputed data is `r meanStepsImpute`.   Median Steps Per Day using Imputed data is `r medianStepsImpute`.      These values are different from the estimates in the first part of the assignment and result in mean and median being equal.


#Are there differences in activity patterns between weekdays and weekends?

```{r addweekday}
par(mfrow = c(2,1), mar= c(4,4,2,1))
weekdayactivity <- subset(activity,weekdays(as.POSIXct(activity$date),abbreviate=TRUE) != "Sun" & weekdays(as.POSIXct(activity$date),abbreviate=TRUE) != "Sat")
StepsPerIntervalWeekday <- with(weekdayactivity, tapply(steps,interval, mean, na.rm=T))
plot(names(StepsPerIntervalWeekday), StepsPerIntervalWeekday, type="l", xlab= "Time of Day", ylab= "Weekday Steps", col="green" , lwd=2)

weekendactivity <- subset(activity,weekdays(as.POSIXct(activity$date),abbreviate=TRUE) == "Sun" | weekdays(as.POSIXct(activity$date),abbreviate=TRUE) == "Sat")
StepsPerIntervalWeekend <- with(weekendactivity, tapply(steps,interval, mean, na.rm=T))
plot(names(StepsPerIntervalWeekend), StepsPerIntervalWeekend, type="l", xlab= "Time of Day", ylab= "Weekend Steps", col="green" , lwd=2)
```
